Automatically Proving Equivalence by Type-Safe Reflection

### 2017-03-26 paper-CICM-17/content/1-introduction.tex
CICM paper : intro currently in progress

Proving that one term is equal to another within some theories is quite common
in formal certification. The two terms are usually not syntactically 
equal---this case would be trivial---but they are equal relative to some
equivalence properties. These properties are either obtained by lemmas
(the implementation of $+$ for natural numbers satisfies the property of
associativity for example), or by axioms.

For example, when working with natural numbers, one might need to prove the
lemma $\forall\ x, y:Nat,\ x*x\ +\ 3*x\ +\ x*y\ =\ x*(3+y+x)$. Or,
when working
with lists and concatenation of lists, one might need to prove the lemma
$\forall\ l1,\ l2,\ l3:List A,\ ((l1\ ++\ (l2 ++ l3))\ ++\ l4\ =\ (l1++l2)\ ++\
(l3++l4)$. Proving this kind of auxiliary lemmas is time consuming,
typically not reusable due to being very specific, and is rarely
the interesting part of formal
certification. Therefore, we would like to be able to automatize as far as
possible the constructions of this kind of proofs. We have implemented
the machinery to do so in Idris, a
dependently typed language~\cite{brady2013idris}.

When doing formal certification, we often require equality proofs between
terms.
For example, say we want to certify the function $f\ :\ A \rightarrow
B$. We have two functions $sem_A\ :\ A \rightarrow SEM$ and $sem_B\ : \ B
\rightarrow SEM$ which associate a semantics to each type A
and B. If this semantics describes all of the important properties of
the behaviour of $f$,
then the lemma of correctness of $f$ can be expressed as $forall\ a:A,\ semA\ a =
semB (f\ a)$. This situation happens very often in model transformations where the behaviour of an object has been preserved by some transformations, as for example in the CompCert certified compiler where the original C program and the generated assembly code are proven to have the same operational semantic \cite{Leroy09}.
Very often, the type $SEM$ has some algebraic
algebraic structure, and it becomes thus possible to
decide equalities in it by following some standard rewriting rules.  
All the equality symbols we have
encountered so far refer to the propositional equality. In the next section,
we will see how this is defined in Idris, and how it differs from the more
primitive judgemental equality.

\subsection{Propositional and Judgemental Equalities in Intentional Type Theories}
In Mathematics, equality is a proposition, e.g. we can disprove an equality or 
assume an equality as a hypothesis.
Since in type theory, propositions are seen as types~\cite{How80}, the
proposition that two elements $x$ and $y$ are equal corresponds to a type.
Thus, if $x$ and $y$ are of type $a$, then the type
$Id_a(x, y)$ represents the proposition ``$x$ is equal to $y$". If this type is
inhabited, then $x$ is said to be provably equal to $y$.  Thus, $Id$ is a type
family (parametrised by the type $a$) indexed over two elements of $a$, giving
$Id (a:Type)\ :\ a \rightarrow a \rightarrow Type$. For
convenience, we write $(Id_a\ x\ y)$ as $(x\ =_a\ y)$. This equality is the
equality which can be manipulated in the language.

There is another, more
primitive, notion of equality in Intentional Type Theories, called judgemental
equality, or definitional equality. This second equality means ``equal by
definition'' The judgemental equality cannot be negated or
assumed; we cannot talk about this primitive equality inside the theory.
Whether or not two expressions are equal by definition is a matter of
evaluating the definitions. For example, if $f\ :\ \mathbb{N}\ \rightarrow\
\mathbb{N}$ is defined by $f\ x\ \equiv\ x\ +\ 2$, then $f\ 5$ is
definitionally equal to $7$. Definitional equality entails unfolding
of functions and reductions, until no more reduction can be
performed. We denote the definitional equality by $\equiv$.

The judgemental equality has to be included in the propositional equal, because
what is equal by definition should be provably equal.  This is accomplished by
giving constructors to the type $Id(a,a)$ and nothing when ``$a$ is not $b$".
One way to implement $Id$ in Idris is:

\begin{lstlisting}
data Id : a -> a -> Type where
     Refl : (x : a) -> Id x x
\end{lstlisting}

The only way for $(Id_a\ x\ y)$ to be inhabited is therefore that $x$ should be
(by definition!) $y$. In this case, the constructor $Refl$ helps to create a
proof of this equality : $(Refl\ x)$ is precisely the proof which says that
$x=_ax$.

When reading these definitions, one could think that the
propositional equality is a simple wrapper of the judgemental equality. This is
not the case: the propositional equality does not only contain the judgemental
equality, because in these type theories, a principle of induction is
associated with each inductive type. This induction principle says that if a
proposition holds for the base cases, and if it can be showed that when it
holds for some terms then it will also hold for the bigger terms obtained by
using the recursive constructors, then this proposition will hold for any term
of this type.

Induction principles work for any proposition, so it also works with the
propositional equality. That means that the axiom of induction principle has
made the type $Id(a,b)$ bigger than it looks like in the above definition, and
there are therefore things which are provably equal which unfortunately aren't
equal syntactically after an expanding of all the definitions. The activity of
proving equalities is therefore in these theories something which isn't
automatically decidable by the type-checker in the general case.

For example, we can prove that $n+0 = n$ by induction, even if $n+0 \not\equiv
n$ with the usual definition of $+$, recursive on its first argument.

These proofs are usually omitted ``on paper'' for everyday mathematics, but
they are required by proof assistants, no matter how obvious they are to the
human reader.
Currently, in the Idris programming language, proofs must be completed
by hand, and consist
of a potentially long sequence of rewriting, where each step of
rewriting uses one of the available properties. These proofs can be
an everyday
routine when working in a dependently typed language, and especially
when intensively indexing types over values in order to capture certain logical
properties.  For example, If a type $T$ is indexed over natural numbers, then
the definition of a function which has codomain $T$ will often require a
proof that the index of the produced term is equal to the expected index. This
proof may use properties of natural numbers and their usual operations:
symmetry and commutativity for +, existence of neutral element for +, etc.

>>>>

Proofs assistants like Coq(ref), Agda(ref) and Idris(ref) are based on Intensional Type Theories, that contain two notions of equality. One is the equality that can be manipulated in the language, called the propositional equality. This equality corresponds to the equality found in mathematics : this is a proposition that can be assumed, negated, proved or disproved. Since in type theory, propositions are seen as types~\cite{How80}, the
proposition that two elements $x$ and $y$ are equal corresponds to a type.
If $x$ and $y$ are of type $a$, then the type
$Id_a(x, y)$ represents the proposition ``$x$ is equal to $y$". If this type is
inhabited, then $x$ is said to be provably equal to $y$. Thus, $Id$ is a type family (parametrised by the type $a$) indexed over two elements of $a$, giving $Id (a:Type)\ :\ a \rightarrow a \rightarrow Type$. For
convenience, we write $(Id_a\ x\ y)$ as $(x\ =_a\ y)$.

The second notion of equality in Intentional Type Theories is a primitive concept, called judgemental equality, or definitional equality. This second equality means ``equal by definition''. The judgemental equality cannot be negated or assumed; we cannot talk about this primitive equality inside the theory.
Whether or not two expressions are equal by definition is a matter of
evaluating the definitions. For example, if $f\ :\ \mathbb{N}\ \rightarrow\
\mathbb{N}$ is defined by $f\ x\ :=\ x\ +\ 2$, then $f\ 5$ is
definitionally equal to $7$. Definitional equality entails unfolding
of functions and reductions, until no more reduction can be
performed. We denote the definitional equality by $\equiv$.

The judgemental equality has to be included in the propositional equality, because what is equal by definition should be provably equal. This is accomplished by giving constructors to the type $Id(a,a)$ and nothing when ``$a$ is not $b$".
In these theories, $Id$ is therefore implemented with the following sum type that has only one constructor :

\begin{lstlisting}
data Id : a $\rightarrow$ a $\rightarrow$ Type where
     Refl : (x : a) $\rightarrow$ Id x x
\end{lstlisting}

The only way for $(Id_a\ x\ y)$ to be inhabited is therefore that $x$ should be
(by definition) $y$. In this case, the constructor $Refl$ helps to create a
proof of this equality : $(Refl\ x)$ is precisely the proof which says that
$x=_ax$.

When reading these definitions, one could wrongly think that the propositional equality is a simple wrapper of the judgemental equality. This is not the case : the propositional equality does not only contain the judgemental equality, because in these type theories, a principle of induction is associated with each inductive type. An inductive principle says that if a proposition holds for the base cases (ie, the constant constructors), and if it can be showed that when it holds for some terms then it will also hold for the bigger terms obtained by using the recursive constructors, then this proposition will hold for any term of this type. More formally, if a type $T$ is an inductive type with a constant constructor and a recursive constructor, ie, $T = 1 + T$ in type-theory, defined in idris as :

\begin{lstlisting}
data T : Type where
  T0 : T
  T1 : T $\rightarrow$ T          
\end{lstlisting}

then we have : \\
$T\_ind : \forall P:T \rightarrow Type,\ (P\ T0) \rightarrow (\forall t:T,\ P\ t \rightarrow P\ (T1\ t)) \rightarrow (\forall t:T,\ P\ t)$.

The principles of induction can be used for proving any proposition, so they also work for the propositional equality. For example, we can prove that $n+0 = n$ for all $n$ by induction on the $Nat\ n$, even if $n+0 \not\equiv n$ with the usual definition of $+$, recursive on its first argument. That means that the axiom of induction has made the type $Id_{a}(x,y)$ bigger than it looks like : it contains terms that are not generated by $Refl$ (the only constructor of $Id_{a}(x,y)$), but that are added by the inductive principles. There are therefore things which are provably equal which unfortunately are not definitionally equal after unfolding all the definitions and after full reductions. The activity of proving equalities is therefore in these theories something which isn't automatically decidable by the type-checker in the very general case.

Proving that one term is equal to another is quite common
in formal certification. For example, when working with natural numbers, one might need to prove the lemma $\forall\ x, y:Nat,\ x*x + 3*x + x*y\ =\ x*(3+y+x)$. Or, when working with lists and concatenation of lists, one might need to prove the lemma $\forall\ l1,\ l2,\ l3:List A,\ (l1 ++ (l2 ++ l3)) ++ l4\ =\ (l1++l2) ++ (l3++l4)$. Currently, in the Idris programming language, these proofs must be completed by hand, and consist of a potentially long sequence of rewriting, where each step of rewriting uses one of the available properties. Proving this kind of auxiliary lemmas is time consuming and typically not reusable due to being very specific. These proofs are usually omitted ``on paper'' for everyday mathematics, but they are required by proof assistants, no matter how obvious they are to the human reader. These proofs can be an everyday routine when working in a dependently typed language, and especially when intensively indexing types over values in order to capture some logical properties, as we will show in the following subsection.
